Live Functional Programming with Typed Holes

### 2018-04-14 calculus.tex
done with pass over Section 3

\matt{Next paragraph can be dropped entirely for space}
%
This approach was originally developed in the reduction semantics of \citet{DBLP:journals/tcs/FelleisenH92} and is the predominant style of operational semantics in the literature on gradual typing. 
Because we distinguish final forms judgementally, rather than syntactically, we use a judgemental formulation of this approach called a \emph{contextual dynamics} by \citet{pfpl}. 
It would be straightforward to derive an equivalent structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} by using search rules instead of evaluation contexts (\citet{pfpl} relates the two approaches).
\todo{Put the search rules in the appendix?}{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% MATT is here (2018.04.14 -- ~7am MT)

@@

First, the expression in function position might be a cast between
arrow types, in which case we apply the arrow cast conversion rule,
Rule {ITApCast}, to rewrite the application form, obtaining an
equivalent application where the expression under the function
cast~$d_1$ is exposed.
%
We know from inverting the typing rules that~$d_1$ has type
$\tarr{\htau_1}{\htau_2}$, and that~$d_2$ has type~$\htau_1'$, where
$\tconsistent{\htau_1}{\htau_1'}$; consequently, we maintain type
safety by placing a cast on~$d_2$ from~$\htau_1'$ to~$\htau_1$.
%
The result of this application will has type $\htau_2$, but the
original cast promised that the result would have consistent type
$\htau_2'$, so we also need a cast on the result from $\htau_2$ to
$\htau_2'$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% MATT is here (2018.04.14 -- ~noon MT)

\subsubsection{Complete Programs} 
%
Although this paper focuses running incomplete programs, it helps
to know that its machinery does not interfere
with reasoning about the behavior of complete programs, i.e. those
with no holes.
%
Appendix~\ref{sec:additional-defns} defines complete programs with the judgements $\isComplete{e}$ and $\isComplete{d}$. 
Failed casts cannot appear in complete internal expressions. 
The following theorem establishes that expansion preserves program completeness.
\begin{thm}[Complete Expansion] ~
  \begin{enumerate}[nolistsep]
    \item
      If $\isComplete{\hexp}$
      and $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\isComplete{\dexp}$.
    \item
      If $\isComplete{\hexp}$
      and $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\isComplete{\dexp}$.
  \end{enumerate}
\end{thm}

@@

The following progress theorem establishes that only classic values result from evaluating a complete program. There are no boxed values or indeterminate forms. 
\begin{thm}[Complete Progress]
  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
  and $\isComplete{\dexp}$
  then either $\stepsToD{}{\dexp}{\dexp'}$
  or $\isValue{\dexp}$.
\end{thm}

@@

The supplemental material includes our mechanization of the semantics and metatheory of \HazelnutLive, 
including proofs of all of the theorems stated above and all necessary lemmas\todo{make this so ian!}{}. We chose the Agda proof assistant \cite{norell2009dependently,norell:thesis} 
(as did the mechanization of \Hazelnut by \citet{popl-paper}, though only a few definitions are common). 
Agda is a good choice because it is designed to explicitly communicate a proof's structure, as is our goal, rather than relying on proof automation. Agda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf. Sec.~\ref{sec:intro}). 
The basic approach is standard: we model judgements as 
inductive datatypes, and rules as dependently typed constructors of these judgements\todo{cite somtehing?}{}. 
We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and encode typing 
contexts, hole contexts and hole environments using metafunctions.\todo{say something about axioms?}{}

@@

The supplemental material also includes a browser-based implementation of \HazelnutLive. 
The implementation is a functional reactive program \cite{DBLP:conf/pldi/CzaplickiC13} 
written using the Reason toolchain for OCaml \cite{reason-what,leroy03:_ocaml} 
together with the OCaml \lismall{React} library \cite{OcamlReact} 
and the \lismall{js_of_ocaml} compiler and its associated libraries \cite{vouillon2014bytecode}. 
Appendix~\ref{sec:ref-impl-screenshots}\todo{do this}{} provides a screenshot of the user interface, which implements all of the live programming features described in Sec.~\ref{sec:examples}, but for the more austere language of this section.

The editor component of the implementation is derived from \Hazelnut. 
It exposes a language of structured edit actions (summarized in the sidebar on the left of Fig.~\ref{fig:screenshot}) that insert holes automatically to guarantee that every edit state has some (possibly incomplete) type. 
Formally,
this corresponds to the Sensibility property from the prior work \cite{popl-paper}.\todo{put statement in the appendix}{} 
By composing this property with the Expandability, Typed Expansion and Progress properties from this paper, we establish a uniquely powerful 
invariant: that every edit state is both statically and dynamically meaningful. 
In other words, live feedback does not ``flicker in and out'' or reflect past edit states. 
The primary purpose of the included implementation is to serve as a proof-of-concept of this invariant, and to be of use to researchers studying the calculus as presented in this section.
% Consistent with this goal, it closely follows the theoretical account in this section, rather than including advanced language features.

We are also separately integrating the mechanisms developed in this paper into the \Hazel programming environment being developed by \citet{HazelnutSNAPL} with the goal of realizing a full-scale hole-driven development experience rooted in the type theoretic foundations established by \Hazelnut. 
Ongoing development is focused on extending the language of \Hazelnut following, in large part, the design of the popular Elm language \cite{Elm,czaplicki2012elm}, and on adding various syntactic and edit-time conveniences that are orthogonal to the topics of this paper. 
For the sake of exposition, the examples in Sec.~\ref{sec:examples} assumed some of these orthogonal features and conveniences, all of which are available in Elm and standard in ML-like languages.
We plan to keep our interpreter for \Hazel, which is an extension of the interpreter in the included implementation, up-to-date as \Hazel evolves.%
\todo{say that we'll use existing results from gradual typing to handle polymorphism and so on}{} 
The included implementation provides a link to \Hazel for interested readers.\todo{bad idea?}{}

Sec.~\ref{sec:discussion} discusses integration of the ideas from \HazelnutLive into other programming systems.\todo{do this}

>>>>

%% \matt{Next paragraph can be dropped entirely for space}
%% %
%% This approach was originally developed in the reduction semantics of \citet{DBLP:journals/tcs/FelleisenH92} and is the predominant style of operational semantics in the literature on gradual typing. 
%% Because we distinguish final forms judgementally, rather than syntactically, we use a judgemental formulation of this approach called a \emph{contextual dynamics} by \citet{pfpl}. 
%% It would be straightforward to derive an equivalent structural operational semantics \cite{DBLP:journals/jlp/Plotkin04a} by using search rules instead of evaluation contexts (\citet{pfpl} relates the two approaches).
%% \todo{Put the search rules in the appendix?}{}

@@

First, the expression in function position might be a cast between
arrow types, in which case we apply the arrow cast conversion rule,
Rule {ITApCast}, to rewrite the application form, obtaining an
equivalent application where the expression~$d_1$ under the function
cast is exposed.
%
We know from inverting the typing rules that~$d_1$ has type
$\tarr{\htau_1}{\htau_2}$, and that~$d_2$ has type~$\htau_1'$, where
$\tconsistent{\htau_1}{\htau_1'}$; consequently, we maintain type
safety by placing a cast on~$d_2$ from~$\htau_1'$ to~$\htau_1$.
%
The result of this application will has type $\htau_2$, but the
original cast promised that the result would have consistent type
$\htau_2'$, so we also need a cast on the result from $\htau_2$ to
$\htau_2'$.

@@

\subsubsection{Complete Programs} 
%
Although this paper focuses running \emph{incomplete} programs, it helps
to know that its machinery does not interfere
with reasoning about the behavior of \emph{complete} programs, i.e. those
with no holes.
%
Appendix~\ref{sec:additional-defns} defines complete programs via judgements~$\isComplete{e}$ and~$\isComplete{d}$. 
%
Failed casts cannot appear in complete internal expressions. 
%
The following theorem establishes that expansion preserves program completeness.
\begin{thm}[Complete Expansion] ~
  \begin{enumerate}[nolistsep]
    \item
      If $\isComplete{\hexp}$
      and $\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}$
      then $\isComplete{\dexp}$.
    \item
      If $\isComplete{\hexp}$
      and $\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}$
      then $\isComplete{\dexp}$.
  \end{enumerate}
\end{thm}

@@

The following progress theorem establishes that evaluating a complete
program always results in classic values, not boxed values nor
indeterminate forms.
%
\begin{thm}[Complete Progress]
  If $\hasType{\hDelta}{\emptyset}{\dexp}{\htau}$
  and $\isComplete{\dexp}$
  then either $\stepsToD{}{\dexp}{\dexp'}$
  or $\isValue{\dexp}$.
\end{thm}

@@

The supplemental material includes our mechanization of the semantics and metatheory of \HazelnutLive, 
including proofs of all of the theorems stated above and all necessary lemmas\todo{make this so ian!}{}. 
%
We chose the Agda proof assistant \cite{norell2009dependently,norell:thesis} 
(as did the mechanization of \Hazelnut by \citet{popl-paper}, though only a few definitions are common). 
%
Agda is a good choice because it is designed to explicitly communicate a proof's structure, as is our goal, rather than relying on proof automation. 
%
Agda itself was also an inspiration for this work because it supports holes, albeit in a more limited form than described here (cf. Sec.~\ref{sec:intro}). 
%
The basic approach is standard: we model judgements as 
inductive datatypes, and rules as dependently typed constructors of these judgements\todo{cite somtehing?}{}. 
%
We adopt Barendregt's convention for bound variables \cite{urban,barendregt84:_lambda_calculus} and encode typing 
contexts, hole contexts and hole environments using metafunctions.\todo{say something about axioms?}{}

@@

The supplemental material includes a browser-based implementation
of \HazelnutLive, which consists of a functional reactive program
%
% Functional Reactive Animation (Hudak 1997) is the canonical FRP cite, not Elm, if you want one; we don't need one, I think. --Matt
%
%\cite{DBLP:conf/pldi/CzaplickiC13} 
%
written with the Reason toolchain for OCaml \cite{reason-what,leroy03:_ocaml} 
together with the OCaml \lismall{React} library \cite{OcamlReact} 
and the \lismall{js_of_ocaml} compiler and its associated libraries \cite{vouillon2014bytecode}. 
%
Appendix~\ref{sec:ref-impl-screenshots}\todo{do this}{} provides a screenshot of the user interface, 
which implements the core calculus of this section, 
along with the corresponding live programming features from \Secref{sec:examples}.

To facilitate interactively editing terms, the implementation employs
the editor semantics of~\Hazelnut, exposing a language of structured
edit actions (summarized in the sidebar on the left of
Fig.~\ref{fig:screenshot})
%
As with prior work, this editor semantics enjoys
the \Property{Sensibility} property by automatically inserting holes
to guarantee that every edit state has some (possibly incomplete) type
.\todo{put statement in the appendix}{}

By composing this meta-theoretic property with
the \Property{Expandability}, \Property{Typed Expansion}
and \Property{Progress} properties from this paper, we establish a
uniquely powerful invariant: \emph{Every} edit state has a \emph{static
meaning} (a la~\Hazelnut) and a \emph{dynamic meaning} (a la~\HazelnutLive).
%
In other words, live feedback never ``flickers in and out'' nor reflects ``stale'' edit states.
%
The current implementation serves as a proof-of-concept of this invariant, 
by providing a working prototype for this section's definitions.

%, and provides to be of use to researchers studying the calculus as presented in this section.
% Consistent with this goal, it closely follows the theoretical account in this section, rather than including advanced language features.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% \matt{I'd drop everything else in this sub-section; 
%% it leaves us open to attack from a hostile reviewer 
%% who is cranky that the implementation is not ``complete'' yet; also, we'd save the space}

%% Separately, we are integrating the mechanisms developed in this paper
%% into the \Hazel programming environment being developed
%% by \citet{HazelnutSNAPL} with the goal of realizing a full-scale
%% hole-driven development experience rooted in the type theoretic
%% foundations established by \Hazelnut.
%% %
%% Ongoing development is focused on extending the language of \Hazelnut following, in large part, the design of the popular Elm language \cite{Elm,czaplicki2012elm}, and on adding various syntactic and edit-time conveniences that are orthogonal to the topics of this paper. 
%% For the sake of exposition, the examples in Sec.~\ref{sec:examples} assumed some of these orthogonal features and conveniences, all of which are available in Elm and standard in ML-like languages.
%% We plan to keep our interpreter for \Hazel, which is an extension of the interpreter in the included implementation, up-to-date as \Hazel evolves.%
%% \todo{say that we'll use existing results from gradual typing to handle polymorphism and so on}{} 
%% The included implementation provides a link to \Hazel for interested readers.\todo{bad idea?}{}

%% Sec.~\ref{sec:discussion} discusses integration of the ideas from \HazelnutLive into other programming systems.\todo{do this}
